---
title: "Ford Go Bike"
output:
  html_notebook: default
  pdf_document: default
---

In this notebook I download and unzip the [Ford Go Bike](https://www.fordgobike.com/) [data](https://www.fordgobike.com/system-data). 

```{r}
library(tidyverse)
library(tictoc)
library(ggmap)
library(skimr)
library(lubridate)
library(forcats)
library(kableExtra)
```

Set working directory.

```{r}
setwd("~/GitHub/Stat6620/fordgobike")
```

Download the files.  First one is not zipped, the remaining are zipped.

```{r}
URL <- "https://s3.amazonaws.com/fordgobike-data/2017-fordgobike-tripdata.csv"
download.file(URL, destfile = "./data/2017-fordgobike-tripdata.csv", method="curl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201801-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201801-fordgobike-tripdata.csv.zip", method="curl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201802-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201802-fordgobike-tripdata.csv.zip", method="curl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201803-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201803-fordgobike-tripdata.csv.zip", method="curl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201804-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201804-fordgobike-tripdata.csv.zip", method="curl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201805-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201805-fordgobike-tripdata.csv.zip", method="curl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201806-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201806-fordgobike-tripdata.csv.zip", method="curl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201807-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201807-fordgobike-tripdata.csv.zip", method="curl")
```

Loop over the one value in the url and filename that changes.

```{r, eval=FALSE}

URL <- "https://s3.amazonaws.com/fordgobike-data/2017-fordgobike-tripdata.csv"
download.file(URL, destfile = "./data/2017-fordgobike-tripdata.csv", method="curl")

for (i in 1:7) {
URL <- paste0("https://s3.amazonaws.com/fordgobike-data/20180",i,"-fordgobike-tripdata.csv.zip")
download.file(URL, destfile = paste0("./data/20180",i,"-fordgobike-tripdata.csv.zip"), method="curl")
}
```



Unzip downloaded files.

```{r}
unzip("./data/201801-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201802-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201803-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201804-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201805-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201806-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201807-fordgobike-tripdata.csv.zip",exdir="./data")
```

Clean up data directory.

```{r}
fn <- "./data/201801-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201802-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201803-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201804-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201805-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201806-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201807-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
```

Read the.csv files

```{r message=FALSE}
fordgobike2017 <- read_csv(file="./data/2017-fordgobike-tripdata.csv")
fordgobike201801 <- read_csv(file="./data/201801-fordgobike-tripdata.csv")
fordgobike201802 <- read_csv(file="./data/201802-fordgobike-tripdata.csv")
fordgobike201803 <- read_csv(file="./data/201803-fordgobike-tripdata.csv")
fordgobike201804 <- read_csv(file="./data/201804-fordgobike-tripdata.csv")
fordgobike201805 <- read_csv(file="./data/201805-fordgobike-tripdata.csv")
fordgobike201806 <- read_csv(file="./data/201806-fordgobike-tripdata.csv")
fordgobike201807 <- read_csv(file="./data/201807-fordgobike-tripdata.csv")
```

Check the head() and tail() of the data.frames that are loaded.


```{r}
head(fordgobike2017) 
head(fordgobike201801)
head(fordgobike201802)
head(fordgobike201803)
head(fordgobike201804)
head(fordgobike201805)
head(fordgobike201806)
head(fordgobike201807)
```



```{r}
tail(fordgobike2017)
tail(fordgobike201801)
tail(fordgobike201802)
tail(fordgobike201803)
tail(fordgobike201804)
tail(fordgobike201805)
tail(fordgobike201806)
tail(fordgobike201807)
```


```{r}
dim(fordgobike2017)

fordgobike2017 %>% count()

```


```{r}

fordgobike201801 %>% count()
fordgobike201802 %>% count()
fordgobike201803 %>% count()
fordgobike201804 %>% count()
fordgobike201805 %>% count()
fordgobike201806 %>% count()
fordgobike201807 %>% count()

```


```{r}
glimpse(fordgobike201805)
glimpse(fordgobike201806)
glimpse(fordgobike201807)


fordgobike201806 <- fordgobike201806 %>%
        mutate(start_station_id = as.integer(start_station_id),
               end_station_id= as.integer(start_station_id) )

fordgobike201807 <- fordgobike201807 %>%
        mutate(start_station_id = as.integer(start_station_id),
               end_station_id= as.integer(start_station_id) )

fordgobike2018 <- bind_rows(fordgobike201801, fordgobike201802, fordgobike201803, fordgobike201804,
                            fordgobike201805, fordgobike201806, fordgobike201807)

glimpse(fordgobike2018)

```



```{r}
fordgobike2018 %>% select(start_station_id,start_station_name, start_station_latitude,start_station_longitude) %>%
  arrange(start_station_id) %>%
  distinct() %>%
  head() %>%
  kable(, format = "rst")
```




```{r}
dim(fordgobike2017)

fordgobike2017 %>% count()

nrow(fordgobike201801) + nrow(fordgobike201802) + nrow(fordgobike201803) + nrow(fordgobike201804)

dim(fordgobike2018)

fordgobike2018 %>% count()

fordgobike <- bind_rows(fordgobike2017, fordgobike2018)

dim(fordgobike)


fordgobike %>% count()

dim(fordgobike)

fordgobike <- fordgobike %>% mutate(age = 2018 - member_birth_year)

fordgobike %>% count() 

dim(fordgobike)

fordgobike <- fordgobike %>% mutate(year=year(start_time), month=month(start_time), day=day(start_time) )

fordgobike %>% count() 

dim(fordgobike)

fordgobike <- fordgobike %>% mutate(week_day = wday(start_time) )

levels <- c("M","T","W","TH","F","SAT","SUN")

fordgobike$week_day <- factor(fordgobike$week_day, levels = levels)

fordgobike %>% count() 

dim(fordgobike)

```



```{r}

today()

now()
```

Age

```{r warning=FALSE}
fordgobike %>% group_by( age ) %>% count()

fordgobike %>% group_by( age ) %>% summary()

skim(fordgobike)

fordgobike %>% ggplot(aes(x=age)) + geom_histogram()

fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age)) + geom_histogram()

fordgobike %>% filter(age <= 100) %>% ggplot(aes(x=age)) + geom_histogram()

fordgobike %>% filter(age > 100) %>% ggplot(aes(x=age)) + geom_histogram()
```


```{r}
fordgobike %>% group_by( member_gender, age ) %>% count()

fordgobike %>% ggplot(aes(x=age, class=member_gender)) + geom_histogram()

fordgobike %>% ggplot(aes(x=age, class=member_gender)) + geom_histogram(aes(y=..density..))

fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age)) + geom_histogram()
```

```{r}
fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age, color=member_gender)) + 
  geom_histogram(position="identity") + 
  facet_grid(member_gender ~ .)

fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age, color=member_gender)) + 
  geom_histogram(aes(y=..density..),position="identity") + 
  facet_grid(member_gender ~ .)

```

Year and day of week.

```{r}

fordgobike %>% ggplot(aes(x=year)) + geom_bar()
fordgobike %>% ggplot(aes(x=month)) + geom_bar() + facet_grid(year ~ .)
fordgobike %>% ggplot(aes(x=day)) + geom_bar() + facet_grid(year ~ .)


```

```{r}
fordgobike2018 <- fordgobike2018 %>% filter(start_station_latitude < 38 & start_station_longitude < 120 )

fordgobike_subset <- fordgobike2018 %>% select(start_station_longitude,start_station_latitude)

fordgobike_subset %>% ggplot(aes(x=start_station_longitude, y=start_station_latitude)) +
  geom_point()
```





```{r}
library(biganalytics)

# run in parallel, the doMC package runs on Windows
library(doParallel)
registerDoParallel(cores = 8)

head(fordgobike2018)

fordgobike_subset2 <- as.matrix(fordgobike_subset)

set.seed <- 123454321

tic()

cl <- bigkmeans(fordgobike_subset2, 3, nstart=8)

toc()

head(cl$cluster)

cl$centers

fordgobike_subset %>% ggplot(aes(x=start_station_longitude, y=start_station_latitude, color=cl$cluster)) +
  geom_point()

fordgobike2018 <- fordgobike2018 %>% mutate(clust = cl$cluster)

```


```{r}
# City of Oakland  c(-122.2711, 37.8044) )

# https://stackoverflow.com/questions/20621250/simple-approach-to-assigning-clusters-for-new-data-after-k-means-clustering

cl$centers

closest.cluster <- function(x) {
  cluster.dist <- apply(cl$centers, 1, function(y) sqrt(sum((x-y)^2)))
  return(which.min(cluster.dist)[1])
}

oak <- closest.cluster(c(-122.2711, 37.8044))
oak

oakland <- fordgobike2018 %>% filter(clust == oak)

oakland %>% ggplot(aes(x=start_station_longitude, y=start_station_latitude)) +
  geom_point()
```



```{r}
tic()

cl.km <- kmeans(fordgobike_subset, 3)

toc()

cl.km$centers

# City of Oakland  c(-122.2711, 37.8044) )

fordgobike_subset %>% ggplot(aes(x=start_station_longitude, y=start_station_latitude, color=cl.km$cluster)) +
  geom_point()
  

```

```{r}
dim(cl$centers)

bayarea <- get_map(location = c(lon=cl$centers[oak,1], lat=cl$centers[oak,2]), zoom = 12, maptype = "roadmap")

ggmap(bayarea)

ggmap(bayarea) +
  geom_point(data = oakland, aes(x = start_station_longitude, y = start_station_latitude), size = 0.2, shape = 19) +
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()
        )

ggsave(filename="oakland.jpg", width = 4, height = 4, units = "cm", plot=last_plot())

ggmap(bayarea) +
  geom_point(data = oakland, aes(x = start_station_longitude, y = start_station_latitude), size = 1, shape = 19) +
  ggtitle("Oakland Ford Go Bike stations")
```



```{r}
bayarea <- get_map(location = "hayward")

ggmap(bayarea)

ggmap(bayarea) +
  geom_point(data = fordgobike2018, aes(x = start_station_longitude, y = start_station_latitude, color  = clust, alpha = 0.1), size = 0.2, shape = 19) +
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        legend.position="none")

ggsave(filename="bayarea.jpg", width = 4, height = 4, units = "cm", plot=last_plot())

ggmap(bayarea) +
  geom_point(data = fordgobike2018, aes(x = start_station_longitude, y = start_station_latitude, color  = clust, alpha = 0.1), size = 1, shape = 19) +
  ggtitle("Bay Area Ford Go Bike stations")
```


Gender of users

```{r}

fordgobike2018 %>% ggplot(aes(x=member_gender, y=duration_sec)) + geom_bar(stat="Identity") +
  ggtitle("Bay Area")

oakland %>% ggplot(aes(x=member_gender, y=duration_sec)) + geom_bar(stat="Identity") +
  ggtitle("Oakland")
```

Duration of rides in the Bay Area

```{r}
fordgobike2018 %>% ggplot(aes(x=duration_sec, y=..density..)) + 
  scale_x_continuous(limits = c(0, 10000)) +
  geom_histogram() +
  geom_density(aes(y=..density..)) 

fordgobike2018 %>% ggplot(aes(log(x=duration_sec), y=..density..)) + 
  geom_histogram() +
  geom_density(aes(y=..density..)) 

fordgobike2018 %>% ggplot(aes(x=duration_sec, y=..density..)) + 
  scale_x_continuous(limits = c(0, 10000)) +
  geom_histogram() + 
  geom_density(aes(y=..density..)) +
  facet_grid(member_gender ~ .)

fordgobike2018 %>% ggplot(aes(log(x=duration_sec), y=..density..)) + 
  geom_histogram() + 
  geom_density(aes(y=..density..)) +
  facet_grid(member_gender ~ .)
```

Durations of rides in Oakland


```{r}
oakland %>% ggplot(aes(x=duration_sec, y=..density..)) + 
  scale_x_continuous(limits = c(0, 10000)) +
  geom_histogram() +
  geom_density(aes(y=..density..)) 

oakland %>% ggplot(aes(log(x=duration_sec), y=..density..)) + 
  geom_histogram() +
  geom_density(aes(y=..density..)) 

oakland %>% ggplot(aes(x=duration_sec, y=..density..)) + 
  scale_x_continuous(limits = c(0, 10000)) +
  geom_histogram() + 
  geom_density(aes(y=..density..)) +
  facet_grid(member_gender ~ .)

oakland %>% ggplot(aes(log(x=duration_sec), y=..density..)) + 
  geom_histogram() + 
  geom_density(aes(y=..density..)) +
  facet_grid(member_gender ~ .)
```

Duration by City

```{r}
fordgobike2018 %>% ggplot(aes(x=duration_sec, y=..density..)) + 
  scale_x_continuous(limits = c(0, 10000)) +
  geom_histogram() +
  geom_density(aes(y=..density..)) +
  facet_grid(clust ~ .) 

fordgobike2018 %>% ggplot(aes(log(x=duration_sec), y=..density..)) + 
  geom_histogram() +
  geom_density(aes(y=..density..)) +
  facet_grid(clust ~ .) 

fordgobike2018 %>% ggplot(aes(x=duration_sec, y=..density..)) + 
  scale_x_continuous(limits = c(0, 10000)) +
  geom_histogram() + 
  geom_density(aes(y=..density..)) +
  facet_grid(clust ~ .) 

fordgobike2018 %>% ggplot(aes(log(x=duration_sec), y=..density..)) + 
  geom_histogram() + 
  geom_density(aes(y=..density..)) +
  facet_grid(clust ~ .) 
```

Duration in Oakland

```{r}
oakland %>% ggplot(aes(x=duration_sec, y=..density..)) + 
  scale_x_continuous(limits = c(0, 10000)) +
  geom_histogram() +
  geom_density(aes(y=..density..)) +
  facet_grid(clust ~ .) 

oakland %>% ggplot(aes(log(x=duration_sec), y=..density..)) + 
  geom_histogram() +
  geom_density(aes(y=..density..)) +
  facet_grid(clust ~ .) 

oakland %>% ggplot(aes(x=duration_sec, y=..density..)) + 
  scale_x_continuous(limits = c(0, 10000)) +
  geom_histogram() + 
  geom_density(aes(y=..density..)) +
  facet_grid(clust ~ .) 

oakland %>% ggplot(aes(log(x=duration_sec), y=..density..)) + 
  geom_histogram() + 
  geom_density(aes(y=..density..)) +
  facet_grid(clust ~ .) 
```



```{r}

fordgobike2018 %>% filter(clust == 1) %>% 
  group_by( member_gender ) %>%
  summarize(dur_mean = mean(duration_sec), dur_sd = sd(duration_sec)) %>%
  kable(format = "rst")


fordgobike2018 %>% filter(clust == 2) %>% 
  group_by( member_gender ) %>%
  summarize(dur_mean = mean(duration_sec), dur_sd = sd(duration_sec)) %>%
  kable(format = "rst")


fordgobike2018 %>% filter(clust == 3) %>% 
  group_by( member_gender ) %>%
  summarize(dur_mean = mean(duration_sec), dur_sd = sd(duration_sec)) %>%
  kable(format = "rst")

```


```{r}
oakland %>%
  group_by( member_gender ) %>%
  summarize(dur_mean = mean(duration_sec), dur_sd = sd(duration_sec)) %>%
  kable(format = "rst")

```


